---
title: "DIA LiP-MS Analysis:  `r params$title`"
subtitle: "`r params$subtitle`"
author: "`r params$author`"
format: 
  html:
    fig-width: 12
    fig-height: 10
    toc: true
    toc-title: 'DIA-LiP Analysis'
    title-block-banner: '#1b2944'
    title-block-banner-color: '#f79646'
    lightbox: auto
execute: 
  echo: false
  warning: false
embed-resources: true
theme: 
 - layout/brand 
 - layout/custom.scss

params:
 description: ""
 input_file_tc: ""
 input_file_lip: ""
 design_file: ""
 fasta_file: ""
 input_file: ""
 folder_prj: ""
 title: ""
 subtitle: ""
 author: "" 
 formula :  ""
 comparisons : ""
 FC_thr : ""
 comparison_label : ""
 complete_rep: ""
 lip_rep: ""
 res_DE: ""
 adjPval_thr: ""
 poi: ""
---

```{r input}
 #| output: false

 required_packages = c('QFeatures',"ggrepel","tidyr","tibble","heatmaply","pcaMethods",
                       "plotly","DT","MSnbase",'gridExtra','dplyr','stringr' ,"ggcorrplot",'logger','pander','ggsci','GGally','yaml')
# #print(required_packages)
 dialipar::check_dependencies(required_packages)

```

## Experiment Description

`r params$description`

```{r check_parameter}
# 
# log_threshold(INFO)
log_appender(appender_console)
log_appender(appender_file(file.path( params$folder_prj, "logfile_dialipa.log")), index = 2)
# formatter_data_frame <- function(df, ...) {
#      pander::pander_return(df, style = 'simple')
# }
# log_formatter(formatter_data_frame)
 log_formatter(formatter_glue)


check_and_substitute_forbidden_chars <- function(input_string) {
  # Define forbidden characters for filenames in Windows
  forbidden_chars <- c("<", ">", ":", "\"", "/", "\\", "|", "?", "*")

  # Iterate over each forbidden character and replace it with an empty space
  for (char in forbidden_chars) {
    input_string <- gsub(char, " ", input_string, fixed = TRUE)
  }
  return(input_string)
}
## present two times , but not exported 
is_empty <- function(x) {
    is.character(x) && length(x) == 1 && x == ''
  }

theme_custom_vis <- function(base_size = 12) {
  theme_bw(base_size = base_size) %+replace%
    theme(
      # leggend
      legend.title = element_text(size = rel(0.85), face = "bold"),
      legend.text = element_text(size = rel(0.70), face = "bold"),
      legend.key = element_rect(fill = "transparent", colour = NA),
      legend.key.size = unit(1.5, "lines"),
      legend.background = element_rect(fill = "transparent", colour = NA),
      strip.background = element_rect(fill = "#7c7c7c", color = "#7c7c7c"),
      strip.text = element_text(size = rel(0.85), face = "bold", color = "#1b2944", margin = margin(5,0,5,0))
    )
}




# make a subfolder for each contrast
if (!dir.exists(file.path(params$folder_prj))){
  dir.create(file.path( params$folder_prj),recursive = TRUE)
}
 dir.create(file.path( params$folder_prj, "Result"),recursive = TRUE)


# theme_set(theme_custom_vis(base_size = 18))
 
st=format(Sys.time(), "%Y%m%d_%H%M")

write(as.yaml(params), file =  file.path( params$folder_prj, paste0("dialipa_parameter_",st  , ".yaml")))

```


```{r import_data}

#L <- readLines(params$design_file, n = 1)
expdesign <- read.delim(params$design_file,sep="\t")
#if (grepl(";", L)) expdesign <- read.csv2(params$design_file) else design <- read.csv(params$design_file)

complete_rep  <-  readRDS(params$complete_rep )

lip_rep  <-  readRDS(params$lip_rep )

res_DE <-  readRDS(params$res_DE )




```


##  General Quality Control

### Identification

```{r identification}
#| lightbox: true


to_plot_a <- complete_rep %>% 
  group_by(Condition, CondRep) %>% 
  summarise(Precursors = length(unique(Precursor.Id)),
            `Protein Groups` = length(unique(Protein.Group))) %>% 
  pivot_longer(-(1:2),
               names_to = "Feature",
               values_to = "IDs") 


qc_idrate <-  ggplot(to_plot_a , aes(x = CondRep, y = IDs, fill = Condition)) +
  geom_col() +
  scale_fill_observable() +
  facet_wrap(~Feature,
             scales = "free_y") +
  labs(title = "Precursor and protein group identificiations per sample",
       x = "Sample",
       y = "Identifications") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90))

qc_idrate

ggsave(   file.path(params$folder_prj, "Result", "IDs.pdf"),plot = qc_idrate,  dpi = 600, height = 6, width = 12)
```

### Abundance Distribution

```{r abundance_distr}
#| layout-ncol: 2
#| lightbox: true


##---------------
to_plot_b  <- complete_rep %>% 
      distinct(CondRep, Precursor.Id, .keep_all = T)


qc_intdistr <- ggplot(to_plot_b, aes(x = CondRep, 
           y = normPQ)) +
  geom_boxplot(aes(fill = Condition)) +
  scale_fill_observable() +
  theme_bw() +
  labs(title = "Normalised precursor quantity distributions per sample",
       x = "Sample",
       y = expression(Log[2](Normalised~Precursor~Quantity))) +
  theme(axis.text.x = element_text(angle = 90))

qc_intdistr

ggsave(   file.path(params$folder_prj, "Result", "intensity_distribution_boxplots.pdf"),plot = qc_idrate,  dpi = 600, height = 6, width = 12)

##-----------------
to_plotc <- complete_rep %>% 
  distinct(CondRep, Precursor.Id, .keep_all = T) 
qc_inthisto <- ggplot(to_plotc , aes(x = normPQ,
           fill = Condition)) +
  geom_histogram() +
  scale_fill_observable() +
  facet_grid(Condition~Replicate) +
  labs(title = "Normalised precursor quantity distributions per sample",
       x = expression(Log[2](Normalised~Precursor~Quantity))) +
  theme_bw()

qc_inthisto

ggsave( file.path(params$folder_prj, "Result","intensity_histogram.pdf"), dpi = 600, height = 6, width = 12)

```

### Missed Cleavage Rate

```{r miscleavage}
#| lightbox: true


MC_alpha_mapping <- c("0" = 1, 
                    "1" = 0.6, 
                    "2" = 0.25)

to_plotd <- complete_rep %>% 
  distinct(CondRep, Precursor.Id, .keep_all = T) 

qc_missclev <-  ggplot(to_plotd, aes(x = Replicate,
           fill = Condition)) +
  geom_bar(aes(alpha = missed_cleavages %>% factor(levels = 4:0)),
           colour = "black") +
  scale_alpha_manual(values = MC_alpha_mapping, guide = guide_legend(override.aes = list(fill = pal_observable()(1)))) +
  scale_fill_observable() +
  facet_wrap(~Condition, 
             scales = "free_x") +
  labs(title = "Missed cleavages per sample",
       x = "Replicate",
       alpha = "Missed cleavages") +
  theme_bw()
qc_missclev

ggsave( file.path(params$folder_prj, "Result","missed_cleavages_rate.pdf"), plot =  qc_missclev , dpi = 600, height = 6, width = 10)


```

###  Pecursor Type Distribution

```{r qc_miss_cleavage}
#| layout-ncol: 2
#| lightbox: true


to_plote <- complete_rep %>% 
    distinct(CondRep, Precursor.Id, pep_type, .keep_all = T) %>% 
  mutate(pep_type = str_replace(pep_type, "(?<!^)T", "-t"))

qc_tryptic <-  ggplot(to_plote , aes(x = Replicate)) +
  geom_bar(aes(fill = pep_type),
           colour = "black"
           ) +
  scale_fill_nejm() +
  facet_wrap(~Condition, 
             scales = "free_x") +
  labs(title = "Peptide types per sample",
       x = "Replicate",
       fill = "Peptide type") +
  theme_bw()

qc_tryptic

ggsave( file.path(params$folder_prj, "Result","peptideType_distribution.pdf"), plot=qc_tryptic, dpi = 600, height = 6, width = 10)


to_plotf <- complete_rep %>% 
    distinct(CondRep, Precursor.Id, pep_type, .keep_all = T) %>% 
    filter(pep_type != "Tryptic") %>% 
    mutate(pep_type = ifelse(pep_type=="NonTryptic",
                            "Non-tryptic",
                            ifelse(AA_last %in% c("R", "K"),
                                    "C-terminal semi-tryptic",
                                    "N-terminal semi-tryptic"))) 

qc_terminality <- ggplot(to_plotf, aes(x = Replicate)) +
  geom_bar(aes(fill = pep_type),
           colour = "black"
           ) +
  scale_fill_nejm() +
  facet_wrap(~Condition, 
             scales = "free") +
  labs(title = "Peptide types per sample excluding tryptic",
       x = "Replicate",
       fill = "Peptide type") +
  theme_bw()

qc_terminality

ggsave( file.path(params$folder_prj, "Result","peptideType_distribution_excluding_tryptic.pdf"), plot=qc_terminality,  dpi = 600, height = 6, width = 10)

```

### Precursor Charge State Distribution

```{r pr_charge_state}
#| lightbox: true

to_plotz <- complete_rep %>% 
  distinct(CondRep, Precursor.Id, pep_type, .keep_all = T) %>% 
  filter(Precursor.Charge<=4) 

qc_chargedistr <- ggplot(to_plotz, aes(x = Replicate)) +
              geom_bar(aes(fill = factor(Precursor.Charge, levels = 4:1)),
                      colour = "black") +
              scale_fill_nejm() +
              facet_wrap(~Condition, 
                        scales = "free_x") +
              labs(title = "Precursor Charge State Distribution",
                  x = "Replicate",
                  fill = "Charge state") +
              theme_bw()

qc_chargedistr

ggsave( file.path(params$folder_prj, "Result","Precursor_Charge_State Distribution.pdf"), plot= qc_chargedistr , dpi = 600, height = 6, width = 10)

```




```{r coverage_qc}

#| layout-ncol: 2
#| lightbox: true

# coverage plot and heatmap) skipped in report 
to_plotj  <- lip_rep %>% 
  distinct(Accession, coverage) 


qc_coverage <- ggplot(to_plotj, aes(x = 100*coverage)) + 
  geom_histogram(bins = 100,
                 alpha = 0.5,
                 colour = "black") +
  scale_x_continuous(limits = c(0, 100),
                     breaks = 0:5*20) +
  labs(title = "Sequence coverage",
       x = "Coverage (%)",
       y = "Proteins") +
  theme_bw()

#qc_coverage

ggsave( file.path(params$folder_prj, "Result","coverage.pdf"), plot= qc_coverage , dpi = 600, height = 6, width = 10)



to_ploth <- complete_rep %>%
            distinct(CondRep, Precursor.Id, .keep_all = T) %>% 
            pivot_wider(id_cols = "Precursor.Id",
                        names_from = "CondRep",
                        values_from = "normPQ") %>% 
            na.omit() %>%
            column_to_rownames("Precursor.Id") %>% 
            log2() %>% 
            cor(method = "spearman")


qc_correlation <-  ggcorrplot(to_ploth) +
  scale_fill_viridis_c() +
  labs(title = "Correlation matrix",
       fill = "Correlation") +
  theme(axis.text.x = element_text(angle = 90))

#qc_correlation

ggsave( file.path(params$folder_prj, "Result","correlationMatrix.pdf"), plot=qc_correlation, dpi = 600, height = 6, width = 10)

```



## Principal Component Analysis 


```{r pca_qc}
pca_input <-
  complete_rep %>%
  distinct(CondRep, Precursor.Id, .keep_all = T) %>% 
  pivot_wider(id_cols = "Precursor.Id",
              names_from = "CondRep",
              values_from = "normPQ") %>% 
  na.omit() %>%
  column_to_rownames("Precursor.Id") %>% 
  # log2() %>% 
  t()

pca_input <- prep(pca_input)
pca_output <- pca(pca_input, method = "svd", nPcs = 3)


pca_df <- scores(pca_output) %>% 
  as.data.frame() %>% 
  rownames_to_column("CondRep") %>%
  left_join(expdesign, by = "CondRep") 



pca_df <- pca_df %>%
  mutate(
    tooltip_text = paste0(
      "Sample: ", CondRep
    )
  )

toplot_pca <- ggplot( pca_df, aes(x = PC1, y = PC2, colour = Condition, text=tooltip_text )) +
  geom_point() +
  scale_colour_observable() +
  # scale_shape_manual(values = c(1, 4, 6)) +
  labs(title = "Precursor-level principal component analysis",
       colour = "Condition",
       x = paste("PC1: ", round(pca_output@R2[1]*100, 1), "% of variance explained", sep = ""),
       y = paste("PC2: ", round(pca_output@R2[2]*100, 1), "% of variance explained", sep = "")
       ) +
  theme_bw()


###
# +
#  geom_text_repel(aes(label = CondRep),
#                  size = 2,
#                  show.legend = F)
##

tosave_pca <- ggplot(pca_df, aes(x = PC1, y = PC2)) +
      geom_point(aes(colour = Condition)) +
      scale_colour_observable() +
      # scale_shape_manual(values = c(1, 4, 6)) +
      labs(title = "Precursor-level principal component analysis",
          colour = "Condition",
          x = paste("PC1: ", round(pca_output@R2[1]*100, 1), "% of variance explained", sep = ""),
          y = paste("PC2: ", round(pca_output@R2[2]*100, 1), "% of variance explained", sep = "")
          ) +
      geom_text_repel(aes(label = CondRep),
                      size = 2,
                      show.legend = F) +
      theme_bw()

ggplotly(toplot_pca)

ggsave(file.path(params$folder_prj, "Result","PCA.pdf"), plot= tosave_pca,  dpi = 600, height = 6, width = 10)


```

## DE Analysis


Using MSqRob2 with the following formula `r params$formula` 

LogFC threshold used  abs(LogFC) >  `r params$FC_thr`; 

Threshold of statistical significance adj.Pval < `r params$adjPval_thr`

`r if (all(params$poi == "")) { "No Proteins of Interest (POIs) are specified; POIs will include accessions with statistically significant peptides" } else {  paste0("Proteins of Interest (POIs) were specified: ", paste(params$poi, collapse = ", "))}`


```{r set_contrast_folder  }
if (!dir.exists(file.path(params$folder_prj, "Result","Contrast"))){
  dir.create(file.path( params$folder_prj, "Result","Contrast"), recursive = TRUE)
}
# old one params$comparisons
for (a in   params$comparison_label ){
  log_info(a)
  log_info(check_and_substitute_forbidden_chars(a))
  dir.create(file.path(params$folder_prj,"Result","Contrast",check_and_substitute_forbidden_chars(a)), recursive = TRUE,showWarnings = TRUE)
}  
log_info({ names(res_DE)})
#log_info('Saving pe object (qfeatures) as Rds file.')
#saveRDS(pe, file = file.path( params$folder_prj, "Result","qfeat_experiment.Rds"))  
```


```{r comparison}
#| results: asis

for (cmp in names(res_DE)) {
        log_info( paste0('  --  ',cmp)) 

    cmp_label <- gsub('Comparison','',cmp )
    cat( '### ', paste(cmp_label,sep=' '), '\n\n')
    cat ('This volcano plot summarizes the differential expression landscape in the comparison between the two groups \n\n')
    #cat(paste('POI : ', paste(res_DE[[cmp]]$POI, collapse = ' ' ),collapse = ' '))
    
    dialipar::render_child(data = res_DE[[cmp]],
                 pe = NULL,
                 label = cmp , 
                 path=file.path(params$folder_prj, "Result","Contrast",check_and_substitute_forbidden_chars(cmp)),  
                 template='_templateContrast.Rmd') 
    cat('\n\n')
    
}
```