---
title: "DIA LiP-MS Analysis:  `r params$title`"
subtitle: "`r params$subtitle`"
author: "`r params$author`"
format: 
  html:
    fig-width: 12
    fig-height: 10
    toc: true
    toc-title: 'DIA-LiP Analysis'
    title-block-banner: '#1b2944'
    title-block-banner-color: '#f79646'
    lightbox: auto
execute: 
  echo: false
  warning: false
embed-resources: true
theme: 
 - layout/brand 
 - layout/custom.scss

params:
 description: ""
 input_file_tc: ""
 input_file_lip: ""
 design_file: ""
 fasta_file: ""
 input_file: ""
 folder_prj: ""
 title: ""
 subtitle: ""
 author: "" 
 formula :  ""
 comparisons : ""
 FC_thr : ""
 comparison_label : ""
 adjPval_thr: ""
 poi: ""
 data_path: ""
---

```{r input}
 #| output: false

 required_packages = c('QFeatures',"ggrepel","tidyr","tibble","heatmaply","pcaMethods",
                       "plotly","DT","MSnbase",'gridExtra','dplyr','stringr' ,"ggcorrplot",'logger','pander','ggsci','GGally','yaml','scater','Glimma')
# #print(required_packages)
 dialipar::check_dependencies(required_packages)

```

## Experiment Description

`r params$description`

```{r check_parameter}
# 
# log_threshold(INFO)
log_appender(appender_console)
log_appender(appender_file(file.path( params$folder_prj, "logfile_dialipa.log")), index = 2)
# formatter_data_frame <- function(df, ...) {
#      pander::pander_return(df, style = 'simple')
# }
# log_formatter(formatter_data_frame)
 log_formatter(formatter_glue)


check_and_substitute_forbidden_chars <- function(input_string) {
  # Define forbidden characters for filenames in Windows
  forbidden_chars <- c("<", ">", ":", "\"", "/", "\\", "|", "?", "*")

  # Iterate over each forbidden character and replace it with an empty space
  for (char in forbidden_chars) {
    input_string <- gsub(char, " ", input_string, fixed = TRUE)
  }
  return(input_string)
}
## present two times , but not exported 
is_empty <- function(x) {
    is.character(x) && length(x) == 1 && x == ''
  }

theme_custom_vis <- function(base_size = 12) {
  theme_bw(base_size = base_size) %+replace%
    theme(
      # leggend
      legend.title = element_text(size = rel(0.85), face = "bold"),
      legend.text = element_text(size = rel(0.70), face = "bold"),
      legend.key = element_rect(fill = "transparent", colour = NA),
      legend.key.size = unit(1.5, "lines"),
      legend.background = element_rect(fill = "transparent", colour = NA),
      strip.background = element_rect(fill = "#7c7c7c", color = "#7c7c7c"),
      strip.text = element_text(size = rel(0.85), face = "bold", color = "#1b2944", margin = margin(5,0,5,0))
    )
}




# make a subfolder for each contrast
if (!dir.exists(file.path(params$folder_prj))){
  dir.create(file.path( params$folder_prj),recursive = TRUE)
}
 dir.create(file.path( params$folder_prj, "Result"),recursive = TRUE)


# theme_set(theme_custom_vis(base_size = 18))
 
st=format(Sys.time(), "%Y%m%d_%H%M")

write(as.yaml(params), file =  file.path( params$folder_prj, paste0("dialipa_parameter_",st  , ".yaml")))

```


```{r import_data}

#L <- readLines(params$design_file, n = 1)
expdesign <- read.delim(params$design_file, sep="\t")
#if (grepl(";", L)) expdesign <- read.csv2(params$design_file) else design <- read.csv(params$design_file)


data_ <- readRDS(params$data_path)
log_info({str(data_, max.level = 2)})

actual_data <- data_$result 

# Now map the variables
complete_rep <- actual_data$qc_data 
pe           <- actual_data$pe
res_DE       <- actual_data$res_DE






```

# General Quality Control 


## QC-1


```{r qc_a}
#| layout-ncol: 2
#| fig-cap: 
#|   - "Precursor and protein group identificiations per sample"
#|   - "Normalised precursor quantity distributions per sample"
#|   - "Normalised precursor quantity distributions per sample (Density)"
#|   - "Normalised precursor quantity distributions per sample (Count)"


ids_ <- complete_rep |>
  select("Condition", "CondRep", "Pipeline","Precursor.Id", "Protein.Group", "value") |>
  filter(!is.na(value)) |>
  group_by(Condition, CondRep) |>
  summarise(Precursors = length(unique(Precursor.Id)),
            `Protein Groups` = length(unique(Protein.Group))) |> 
  tidyr::pivot_longer(-(1:2),
               names_to = "Feature",
               values_to = "IDs") |> 
  ggplot(aes(x = CondRep, y = IDs, fill = Condition)) +
  geom_col() +
  scale_fill_observable() +
  facet_wrap(~Feature,
             scales = "free_y") +
  labs(title = "Precursor and protein group identificiations per sample",
       x = "Sample",
       y = "Identifications") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90))

ids_

ggsave(   file.path(params$folder_prj, "Result", "IDs.pdf"),plot = ids_,  dpi = 600, height = 6, width = 12)

normalized_pre <- complete_rep |>
  select("Condition", "CondRep", "Pipeline","value") |>
  filter(!is.na(value)) |>
  ggplot(aes(x = CondRep, 
             y = value)) + 
    geom_boxplot(aes(fill = Condition)) +
  scale_fill_observable() +
  theme_bw() +
  labs(title = "Normalised precursor quantity distributions per sample",
       x = "Sample",
       y = expression(Log[2](Normalised~Precursor~Quantity))) +
  theme(axis.text.x = element_text(angle = 90))

normalized_pre


ggsave(   file.path(params$folder_prj, "Result", "normalized_int_I.pdf"),plot = normalized_pre,  dpi = 600, height = 6, width = 12)

normalized_preII <- complete_rep |>
  select("value", "CondRep","Treatment","Pipeline") |>
  filter(!is.na(value)) |>
  ggplot() + 
  aes(x = value, 
      color = Treatment, 
      group = CondRep) +
  geom_density() +
facet_wrap(~Pipeline) +
  labs(title = "Normalised precursor quantity distributions per sample (Density)",
       x = expression(Log[2](Normalised~Precursor~Quantity))) +  
  theme_bw()
normalized_preII

ggsave(   file.path(params$folder_prj, "Result", "normalized_int_II.pdf"),plot = normalized_preII,  dpi = 600, height = 6, width = 12)

normalized_preIII <- complete_rep |>
  select("value", "Condition", "Replicate") |>
  filter(!is.na(value)) |>
  ggplot(aes(x = value,
           fill = Condition)) +
  geom_histogram() +
  scale_fill_observable() +
  facet_grid(Condition~Replicate) +
  labs(title = "Normalised precursor quantity distributions per sample",
       x = expression(Log[2](Normalised~Precursor~Quantity))) +
  theme_bw()

normalized_preIII

ggsave(   file.path(params$folder_prj, "Result", "normalized_int_III.pdf"),plot = normalized_preIII
,  dpi = 600, height = 6, width = 12)
```


## QC-2
```{r qc_b}
#| layout-ncol: 2
#| fig-cap: 
#|   - "Peptide types per sample"
#|   - "Non-Tryptic peptide typed per sample"
#|   - "Charge state per sample"
#|   - "Missed cleavages per sample"


trypicity <- complete_rep |>
  filter(assay %in% c("LiP", "TC")) |>
  select("value", "pep_type", "Replicate", "Condition") |>
  filter(!is.na(value)) |>
  mutate(pep_type = str_replace(pep_type, "(?<!^)T", "-t")) |> 
ggplot(aes(x = Replicate)) +
  geom_bar(aes(fill = pep_type),
           colour = "black"
           ) +
  scale_fill_nejm() +
  facet_wrap(~Condition, 
             scales = "free_x") +
  labs(title = "Peptide types per sample",
       x = "Replicate",
       fill = "Peptide type") +
  theme_bw()

trypicity

ggsave(   file.path(params$folder_prj, "Result", "tripticity_peptide.pdf"),plot = trypicity
,  dpi = 600, height = 6, width = 12)


terminally <- complete_rep |>
  select("value","pep_type", "AA_last", "Replicate", "Condition") |> 
  filter(!is.na(value)) |>
  filter(pep_type != "Tryptic") |>
  mutate(pep_type = ifelse(pep_type=="NonTryptic",
                           "Non-tryptic",
                           ifelse(AA_last %in% c("R", "K"),
                                  "C-terminal semi-tryptic",
                                  "N-terminal semi-tryptic"))) |> 
ggplot(aes(x = Replicate)) +
  geom_bar(aes(fill = pep_type),
           colour = "black"
           ) +
  scale_fill_nejm() +
  facet_wrap(~Condition, 
             scales = "free") +
  labs(title = "Peptide types per sample",
       x = "Replicate",
       fill = "Peptide type") +
  theme_bw()

terminally

ggsave(   file.path(params$folder_prj, "Result", "terminally_peptide.pdf"),plot = terminally
,  dpi = 600, height = 6, width = 12)

charge_state <- complete_rep |>
  select("value", "Precursor.Charge", "Replicate", "Condition") |>
  filter(!is.na(value)) |>
  filter(Precursor.Charge<=4) |> 
ggplot(aes(x = Replicate)) +
  geom_bar(aes(fill = factor(Precursor.Charge, levels = 4:1)),
           colour = "black") +
  scale_fill_nejm() +
  facet_wrap(~Condition, 
             scales = "free_x") +
  labs(title = "Charge state per sample",
       x = "Replicate",
       fill = "Charge state") +
  theme_bw()

charge_state
ggsave(   file.path(params$folder_prj, "Result", "charge_state_peptide.pdf"),plot = charge_state
,  dpi = 600, height = 6, width = 12)

MC_alpha_mapping <- c("0" = 1, 
                    "1" = 0.6, 
                    "2" = 0.25)

miss_cl  <- complete_rep|>
  select("value", "Replicate", "Condition", "Stripped.Sequence") |>
  filter(!is.na(value)) |>
    mutate(missed_cleavages = Stripped.Sequence |> 
             str_count("[RK](?!(P|$))")) |>
  ggplot(aes(x = Replicate,
             fill = Condition)) +
  geom_bar(aes(alpha = missed_cleavages |> factor(levels = 4:0)),
           colour = "black") +
  scale_alpha_manual(values = MC_alpha_mapping, guide = guide_legend(override.aes = list(fill = pal_observable()(1)))) +
  scale_fill_observable() +
  facet_wrap(~Condition, 
             scales = "free_x") +
  labs(title = "Missed cleavages per sample",
       x = "Replicate",
       alpha = "Missed cleavages") +
  theme_bw()


miss_cl


ggsave(   file.path(params$folder_prj, "Result", "miss_cleavage_peptide.pdf"),plot = miss_cl
,  dpi = 600, height = 6, width = 12)
```


## PCA / MDS 


```{r }

se <- joinAssays(pe,i=c("precursors_lip_norm","precursors_tc_norm"),  fcol = "Precursor.Id") |>  getWithColData("joinedAssay")

se <- scater::runMDS(as(se, "SingleCellExperiment"), exprs_values =1)

eig <- attr(reducedDim(se, "MDS"), "eig")

# Calculate % variance (Eigenvalue / Sum of all Eigenvalues)
if (!is.null(eig)) {
  percent_var <- (eig / sum(eig)) * 100
  xlbl <- paste0(" Component 1 (", round(percent_var[1], 1), "%)")
  ylbl <- paste0("Component 2 (", round(percent_var[2], 1), "%)")
} else {
  # Fallback if no variance is found
  xlbl <- "MDS1"
  ylbl <- "MDS2"
}


mds_plot <- scater::plotMDS(se, colour_by = "Condition") +
  # We re-apply geom_point to control alpha and the custom hover text
  # Using \n creates a new line in the plotly tooltip
  geom_point(aes(color = colData(se)$Condition, 
    text = paste0(
      "Sample: ", colData(se)$CondRep, "<br>",
      "Condition: ", colData(se)$Condition, "<br>",
      "Replicate: ", colData(se)$Replicate
    )
  ), alpha = 0.8, size = 3) + 
  labs(x = xlbl, y = ylbl) +
  theme_bw()

ggplotly(mds_plot, tooltip = "text")
# mds_coords <- reducedDim(se, "MDS")
# mdata <- as.data.frame(colData(se))


# glimmaMDS(assay(se), 
#           groups = colData(se)$Condition,
#           labels = colData(se)$CondRep,
#           sample.data = mdata)

```



## DE Analysis


Using MSqRob2 with the following formula `r params$formula` 

LogFC threshold used  abs(LogFC) >  `r params$FC_thr`; 

Threshold of statistical significance adj.Pval < `r params$adjPval_thr`

`r if (all(params$poi == "")) { "No Proteins of Interest (POIs) are specified; POIs will include accessions with statistically significant peptides" } else {  paste0("Proteins of Interest (POIs) were specified: ", paste(params$poi, collapse = ", "))}`

```{r set_contrast_folder  }
if (!dir.exists(file.path(params$folder_prj, "Result","Contrast"))){
  dir.create(file.path( params$folder_prj, "Result","Contrast"), recursive = TRUE)
}
# old one params$comparisons
for (a in   params$comparison_label ){
  log_info(a)
  log_info(check_and_substitute_forbidden_chars(a))
  dir.create(file.path(params$folder_prj,"Result","Contrast",check_and_substitute_forbidden_chars(a)), recursive = TRUE,showWarnings = TRUE)
}  

log_info({ names(res_DE)})

log_info('Saving pe object (qfeatures) as Rds file.')
saveRDS(pe, file = file.path( params$folder_prj, "Result","qfeat_experiment.Rds"))  
```


```{r comparison}
#| results: asis

for (cmp in names(res_DE)) {
        log_info( paste0('  --  ',cmp)) 

    cmp_label <- gsub('Comparison','',cmp )
    cat( '### ', paste(cmp_label,sep=' '), '\n\n')
    cat ('This volcano plot summarizes the differential expression landscape in the comparison between the two groups \n\n')
    #cat(paste('POI : ', paste(res_DE[[cmp]]$POI, collapse = ' ' ),collapse = ' '))
    
    dialipar::render_child(data = res_DE[[cmp]],
                 pe = NULL,
                 label = cmp , 
                 path=file.path(params$folder_prj, "Result","Contrast",check_and_substitute_forbidden_chars(cmp)),  
                 template='_templateContrast.Rmd') 
    cat('\n\n')   
}
```